---
title: "Healthcare Cost Analysis"
output: html_notebook
--- 
### Background and Objective:

A nationwide survey of hospital costs conducted by the US Agency for Healthcare consists of hospital records of inpatient samples. The given data is restricted to the city of Wisconsin and relates to patients in the age group 0-17 years. The agency wants to analyze the data to research on healthcare costs and their utilization.

---

### Import Dataset
```{r}
#setting working directory
setwd("D:/Tech_Code/Simplilearn_materials/Simplilearn_R/Simplilearn_R_Projects/R_Simplilearn_project7/Simplilearn_R_projects")
getwd()
```

```{r}
#install.packages("readxl")
library(readxl)

Hospital_costs <- read_excel("hospitalCosts.xlsx")
head(Hospital_costs)
```



### 1. Record patient statistics:
The agency wants to find the age category of people who frequently visit the hospital and has the maximum expenditure.

* **AGE**: Age of the patient discharged
* **TOTCHG**: Hospital discharge costs
```{r}
summary(Hospital_costs)
```

### Get the number hospital visits based on age
```{r}
summary(as.factor(Hospital_costs$AGE))
```
* Total number of hospital visits for 0-1 age group is 307
```{r}
hist(Hospital_costs$AGE, main = "Histogram of AgeGroup vs Hospital visits",
     xlab = "Age Groups",border = "black", col = c("light green","dark green"), xlim=c(0,20), ylim=c(0,350))
```
* As it can be seen here, from the above histogram, the maximum number of hospital visits are for the age group 0-1 years 

### Summarized expenditure based on age group
```{r}
Expense_Age <- aggregate(TOTCHG ~ AGE, FUN = sum, data = Hospital_costs)
```

### Get the maximum expense and its corresponding age group
```{r}
which.max(tapply(Expense_Age$TOTCHG, Expense_Age$TOTCHG, FUN=sum))
barplot(tapply(Expense_Age$TOTCHG, Expense_Age$AGE, FUN=sum),
        col = "light blue")
```
* Maximum expenditure for 0-1 years is 678118



### 2. Diagnosis-related group that has maximum hospitalization and expenditure

In order of severity of the diagnosis and treatments and to find out the expensive treatments, the agency wants to find the diagnosis-related group that has maximum hospitalization and expenditure.

* **APRDRG**: All Patient Refined Diagnosis Related Groups 
            -is a classification system that classifies patients according to their reason of admission, severity of illness and risk of mortality.
* **TOTCHG**: Hospital discharge costs
```{r}
summary(as.factor(Hospital_costs$APRDRG))
```

Get the diagnosis-related group and its hospitalization expenditure
```{r}
DiagnosisCost <- aggregate(TOTCHG ~ APRDRG, FUN = sum, data = Hospital_costs)
```

Get the maximum diagnostic cost
```{r}
DiagnosisCost[which.max(DiagnosisCost$TOTCHG),]
```
* From the above snippet, we can see here that *640* diagnosis related group had maximum cost of *437978*



### 3. Race vs Hospitalization costs

To make sure that there is no malpractise, the agency needs to analyze if the race of the patient is related to the hospitalization costs.

* H0(Null Hypothesis): Independent variable (RACE) is not influencing dependent variable (COSTS)
* H0: There is no correlation among the residuals
* H1: There is a correlation among the residuals
* p-value < 0.5(significant value) ---> reject the null hypothesis
  - (we need high p-value so that we cannot reject the null hypothesis)

```{r}
summary(as.factor(Hospital_costs$RACE))
```
* There is one null value. This need to be removed.
```{r}
Hospital_costs <- na.omit(Hospital_costs)
summary(as.factor(Hospital_costs$RACE))
```
* As it can be seen, 484 patients out of 499 fall under group 1, showing that the number of observations in group 1 is way higher than that of others - hence the data is skewed. This will only affect the results from linear regression or ANOVA analysis.
```{r}
raceInfluenceAOV <- aov(TOTCHG ~ RACE, data = Hospital_costs)
raceInfluenceAOV
```

```{r}
summary(raceInfluenceAOV)
```
* The residual variance(deviation from original)(of all other variables) is very high. 
  i) This implies that there is very little influence from RACE on hospitalization costs.
* Also, it can be seen that the degree of freedom(DF) for RACE is 1 and that of residuals is 497 obs.
* The F-value(test-statistic) is 0.16, which is much lesser than 0.5 showing that the race doesnt affect the hospitalization costs.
* The Pr(>F)(p-value) of 0.69 is high, conforming that RACE does not affect hospitalization costs.



### 4. To properly utilize the costs, the agency has to analyze the severity of the hospital costs by age and gender for the proper allocation of resources.
```{r}
Hospital_costs$FEMALE <- as.factor(Hospital_costs$FEMALE)
summary(Hospital_costs$FEMALE)
```
* There is an equal distribution of male and female in the group.

```{r}
ageGenderInflModel <- lm(TOTCHG ~ AGE+FEMALE, data=Hospital_costs)
summary(ageGenderInflModel)
```
* Since the p-values of AGE is much lesser than 0.05 and also lesser than 0.001, the ideal statistical significance level(***), it means AGE is the most statistically sugnificant.
* Similarly , gender is also less than 0.05.
* Hence we can conclude that the model is statistically significant.



### 5. Since the length of stay is the crucial factor for in-patients, the agency wants to find if the length of stay can be predicted from age, gender, and race. 
```{r}
Hospital_costs$RACE <- as.factor(Hospital_costs$RACE)

ageGenderRaceInflModel=lm(LOS ~ AGE+FEMALE+RACE, data=Hospital_costs)
summary(ageGenderRaceInflModel)
```
* The p-value is higher than 0.05 for age, gender and race, indicating there is no linear relationship between these variables and length of stay.
* Hence age, gender and race cannot be used to predict the length of stay of impatients.



### 6. Complete Analysis
The agency wants to find the variable that mainly affects hospital costs.
**Significance method** - build a model using all independent variables vs dependent variable
```{r}
HospitalcostModel <- lm(TOTCHG ~ ., data=Hospital_costs)
summary(HospitalcostModel)
```
* As it is apparent that, from the coefficient values Age, Length of stay(LOS) and Patient Redefined diagnosis related groups(APRDRG) are the once with the statistical significance.
  - Race is least significant.

### Build a model removing RACE.
```{r}
hospcostM1 = lm(TOTCHG ~ .-RACE, data = Hospital_costs)
summary(hospcostM1)
```
```{r}
hospcostM2 = lm(TOTCHG ~ .-(RACE+FEMALE), data = Hospital_costs)
summary(hospcostM2)
```

* Since APRDRG has -ve t-value, dropping it (negative correlation)
```{r}
hospcostM3 = lm(TOTCHG ~ AGE+LOS, data = Hospital_costs)
summary(hospcostM3)
```

### Comparing Models
* Creating the required columns:
```{r}
Data <- rep("HospitalCosts",4)
Approach <- rep("Ap1:significance",4)
Model_Name <- c("HospitalcostModel", "hospcostM1", "hospcostM2", "hospcostM3")
Detail <- c("signif, all independent variables", "-RACE", "-(RACE+FEMALE(gender))", "AGE+LOS")

R2 <- c(0.554,0.553,0.551,0.419)
adj_R2 <- c(0.549,0.549,0.548,0.416)
std_err <- c(2610, 2610, 2620, 2970)
R2_sub_adjR2 <- c(0.005,0.004,0.003,0.003)
pvalue <- rep("<2e-16", 4)
```

* Structuring the columns of the Dataframe/ Table:
```{r}
ModelTable <- data.frame(Data, Approach, Model_Name, Detail, R2, adj_R2, std_err, R2_sub_adjR2, pvalue)
names(ModelTable)[c(3,6,7,8)] <- c("Model Name","adj R2","std error", "R2 - adj R2") 

ModelTable
```
* Removing Race[RACE] and Gender[FEMALE] doesn’t change the R2 value. It doesn’t impact cost
* Removing APRDRG in model **hospcostM2** increases the standard error. Hence model **hospcostM2** seems to be better.

---
## ANALYSIS CONCLUSIONS

* As it is evident in multiple models drawn above, health care costs is dependent on age, length of stay and type.
   - Health care cost is highest for patients between 0-1 year.
* Maximum expenditure for 0-1 year is *678118* 
* Length of Stay increases the stay charge.
* All Patients Refined Diagnosis Related Groups also affects Health care costs.
   - 640 diagnosis related group had a max cost of 437978.
* Race and gender doesn't have much impact on hospital cost.


